---
title: "Cyclist_Project"
author: "Amit Bhaskar"
date: "2025-01-01"
output: html_document
---

## Introduction

##Business Objective

##Data

```{r}
install.packages("tidyverse")
library("tidyverse")
library(lubridate)
library(ggplot2)

#Reading the dataset CSVs
Jan<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202401-divvy-tripdata.csv")
Feb<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202402-divvy-tripdata.csv")
Mar<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202403-divvy-tripdata.csv")
Apr<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202404-divvy-tripdata.csv")
May<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202405-divvy-tripdata.csv")
Jun<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202406-divvy-tripdata.csv")
Jul<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202407-divvy-tripdata.csv")
Aug<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202408-divvy-tripdata.csv")
Sep<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202409-divvy-tripdata.csv")
Oct<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202410-divvy-tripdata.csv")
Nov<- read_csv("C:/Users/Tnluser/Desktop/Cyclist Project/202411-divvy-tripdata.csv")
```

Now our aim is to consolidate the data of these 11 csv sheets into one big data frame.. but before that we have to do some transformations. 1) All column names must be the same. 2) All columns must be in the same order so as to facilitate proper stacking 3) All columns must have the same respective data types in each sheet.

```{r}
#Inspecting column names of the CSVs
colnames(Jan)
colnames(Feb)
colnames(Mar)
colnames(Apr)
colnames(May)
colnames(Jun)
colnames(Jul)
colnames(Aug)
colnames(Sep)
colnames(Oct)
colnames(Nov)
```

1)  and 2) It turns out that all 11 sheets have the same column names and in the same order as that in Jan 2024 sheet. A total of 13 columns given below

colnames(Jan) [1] "ride_id" "rideable_type" "started_at" "ended_at"\
[5] "start_station_name" "start_station_id" "end_station_name" "end_station_id"\
[9] "start_lat" "start_lng" "end_lat" "end_lng"\
[13] "member_casual"

3)  Now lets check the data types of each column in each sheet

```{r}
str(Jan)
str(Feb)
str(Mar)
str(Apr)
str(May)
str(Jun)
str(Jul)
str(Aug)
str(Sep)
str(Oct)
str(Nov)
```

We have consistent data types across the sheets for respective columns.

1.  ride_id - character
2.  rideable_type - character
3.  started_at - datetime
4.  ended_at - datetime
5.  start_station_name - character
6.  start_station_id - character
7.  end_station_name - character
8.  end_station_id - character
9.  start_lat - double
10. start_lng - double
11. end_lat - double
12. end_lng - double
13. member_casual - character

```{r}
```

Now let's append the data into one big data frame.

```{r}
Trips2024 <- bind_rows(Jan,Feb,Mar,Apr,May,Jun,Jul,Aug,Sep,Oct,Nov)
```

For our purposes, information about latitudes and longitudes doesn't hold that much value, so dropping off these 4 columns. Also we would do well to get the weekday, month in separate columns from the start time column.

```{r}
#Removing start_lat, end_lat, start_lng, end_lng columns
Trips2024 <- Trips2024 %>%
  select(-start_lat, -end_lat, -start_lng, -end_lng)

#Renaming rideable_type column name to "bike_type" and "member_casual" to "customer_type" for better readability
Trips2024 <- Trips2024 %>%
  rename(`bike_type` = rideable_type)
Trips2024 <- Trips2024 %>%
  rename(`customer_type` = member_casual)

#Getting day,month and start_time in separate columns using lubridate
install.packages("lubridate")
library(lubridate)


Trips2024 <- Trips2024 %>%
  mutate(
    day_of_ride = wday(started_at, label = TRUE, abbr = TRUE),  # Extract day of the week
    month = month(started_at, label = TRUE, abbr = TRUE),# Extract month
  )

#Getting duration of rides in minutes

Trips2024 <- Trips2024 %>%
  mutate(duration = round(as.numeric(difftime(ended_at, started_at, units = "mins")), 1))
```

Let's have a quick confirmation of how many unique bike types are there and how many unique customer types.

```{r}
cat("Unique bike types:", n_distinct(Trips2024$`bike-type`), "\n")
cat("Unique customer types:", n_distinct(Trips2024$customer_type), "\n")
```

```{r}
str(Trips2024)
```

```{r}
library(ggplot2)

# Prepare data
ride_share <- Trips2024 %>%
  group_by(customer_type) %>%
  summarise(total_rides = n()) %>%
  mutate(percentage = total_rides / sum(total_rides) * 100)

# Calculate total rides in 2024
total_rides_2024 <- sum(ride_share$total_rides)

# Create the pie chart
ggplot(ride_share, aes(x = "", y = percentage, fill = customer_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +  # Transform bar chart to pie chart
  labs(title = "Percentage Share of Casuals and Members in 2024 Rides",
       x = NULL, y = NULL) +
  theme_void() +  # Removes axis and grid lines
  theme(legend.title = element_blank())+  # Removes legend title
geom_text(aes(label = paste0( total_rides, "\n", 
                               round(percentage, 1), "%")),  # Add ride count and percentage labels
            position = position_stack(vjust = 0.5), size = 5, color = "white")+
  annotate("text", 
           x = 1, y = 0.5, 
           label = paste("Total Rides:", total_rides_2024), 
           hjust = 0.5, vjust = 0, 
           size = 4, color = "black")

```

```{r}
# Prepare the data: count rides by customer type and month
rides_by_month <- Trips2024 %>%
  group_by(customer_type, month) %>%
  summarise(no_of_rides = n(), .groups = "drop")  # Avoids grouping warning

# Create the plot
ggplot(rides_by_month, aes(x = factor(month), y = no_of_rides, fill = customer_type)) +
  geom_bar(stat = "identity", position = "dodge",width = 0.7  ) +
  labs(title = "Number of Rides by Customer Type Across Months",
       x = "Month", y = "Number of Rides") +
  scale_x_discrete(labels = month.abb) + # Use scale_x_discrete() for months
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  theme_minimal() +
  theme(legend.title = element_blank())  # Removes legend title
```

```{r}

# Prepare the data: count rides by customer type and weekday
rides_by_weekday <- Trips2024 %>%
  group_by(customer_type, day_of_ride) %>%
  summarise(no_of_rides = n(), .groups = "drop")  # Avoids grouping warning

# Create the plot
ggplot(rides_by_weekday, aes(x = factor(day_of_ride, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                             y = no_of_rides, fill = customer_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +  # Decrease width for more space between bars
  labs(title = "Number of Rides by Customer Type Across Weekdays",
       x = "Weekday", y = "Number of Rides") +
  scale_x_discrete(labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")) +  # Abbreviated weekday names
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels as commas
  theme_minimal() +
  theme(legend.title = element_blank())+  # Removes legend title
  geom_text(aes(label = no_of_rides), position = position_dodge(width = 0.7), vjust = -0.3, size = 2.9)  # Show number above bars

```

4)Avg duration of rides by causal and members across weekdays

```{r}

# Prepare the data: calculate average duration by customer type and weekday
avg_duration_by_weekday <- Trips2024 %>%
  group_by(customer_type, day_of_ride) %>%
  summarise(avg_duration = round(mean(duration), 1), .groups = "drop")  # Calculate average duration and round to 1 decimal place

# Create the plot
ggplot(avg_duration_by_weekday, aes(x = factor(day_of_ride, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                                    y = avg_duration, fill = customer_type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +  # Bar chart with dodged bars
  labs(title = "Average Duration of Rides by Customer Type Across Weekdays",
       x = "Weekday", y = "Average Duration (minutes)") +
  scale_x_discrete(labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")) +  # Abbreviated weekday names
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels as commas
  theme_minimal() +
  theme(legend.title = element_blank()) +  # Removes legend title
  geom_text(aes(label = avg_duration), position = position_dodge(width = 0.7), vjust = -0.3, size = 3)  # Show average duration above bars

```

5)Ride start times for casuals and members across time of day

```{r}
library(tidyverse)

# Ensure 'started_at' is in the correct datetime format
Trips2024 <- Trips2024 %>%
  mutate(started_at = as.POSIXct(started_at))  # If not already in POSIXct format

# Extract the hour of the day from 'started_at'
Trips2024 <- Trips2024 %>%
  mutate(hour_of_day = format(started_at, "%H"))

# Group by 'customer_type' and 'hour_of_day', then count the number of rides
ride_count_by_time <- Trips2024 %>%
  group_by(customer_type, hour_of_day) %>%
  summarise(ride_count = n(), .groups = 'drop')

# Plot the line graph
ggplot(ride_count_by_time, aes(x = as.numeric(hour_of_day), y = ride_count, color = customer_type)) +
  geom_line() +
  labs(
    title = "Rides Start Time Across Time of Day",
    x = "Hour of Day",
    y = "Number of Rides",
    color = "Customer Type"
  ) +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +  # Prevent scientific notation on y-axis
  theme_minimal()



```

```{r} 
ggplot(data = Trips2024, aes(x = bike_type, fill = customer_type)) +
  geom_bar(position = "dodge", stat = "count") +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    position = position_dodge(width = 0.9),
    vjust = -0.5
  ) +
  labs(
    title = "Number of Rentals by Members and Casuals by Bike Type",
    x = "Bike Type",
    y = "Number of Rentals",
    fill = "Customer Type"
  ) +
  theme_minimal()

```
